Optimal control with phase constraints for a quasilinear endovenous laser ablation model

### 2022-11-02 phd_5/article.tex
nasty and awful translation

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mhchem}
\usepackage{stmaryrd}
\usepackage{bbold}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\graphicspath{ {./images/} }

@@

\begin{abstract}
An optimal control problem for quasilinear equations of radiative-conductive heat transfer simulating the process of endovenous laser ablation in a bounded domain with the reflecting boundary is studied. In a given subdomain, it is required to provide a given temperature distribution, while the temperature in another given subdomain cannot exceed a critical value. On the basis of estimates of the solution of the controlled system, the solvability of the control problem is proved. An algorithm for solving the optimal control problem by minimizing the objective functional with a penalty is proposed. The efficiency of the algorithm is illustrated by a numerical example.
\end{abstract}

\section{INTRODUCTION}
The procedure of endovenous laser ablation (EVLA) is safe and sufficiently effective in the treatment of varicose veins [1]. During EVLA, a laser optical fiber is inserted into the damaged vein. Then the laser radiation is transmitted through the fiber, which at this time is pulled out of the vein. The end of the optical fiber is usually covered with a carbonized layer (optical fiber tip). The carbonized layer divides the laser energy into the optical fiber tip heating and radiation. The heat from the optical fiber tip is transferred to the blood by the conductive heat exchange, the heat transfer is significantly increased by the flow of bubbles formed at the heated fiber tip. The radiation entering the blood and surrounding tissue is partially absorbed with the release of heat. As a result, the generated thermal energy causes significant heating of the vein which leads to its obliteration (vein closure).

Mathematical modeling of radiation and thermal processes arising during EVLA is important to determine optimal parameters of radiation that provide a sufficiently high temperature inside the vein for successful obliteration, on the other hand, the generated temperature field should be relatively safe for the tissue surrounding the vein. The results of numerical simulation of EVLA for wavelengths in the range from 810 to $1470 \mathrm{~nm}$ and different vein diameters are discussed in [2-4]. In [5], the importance of mathematical modeling for choosing the optimal power of laser radiation and pullback speed of the fiber is noted. Usually, when choosing the optimal radiation parameters, the direct multiple modeling is used [2-5]. This approach is not effective in a practical point of view. Moreover, issues related to the existence and uniqueness of the optimal regime remain open.

The most promising approach to choose the optimal parameters of radiation is to consider an optimal control problem for equations of reaction-diffusion type describing the EVLA procedure. Different approaches to the analysis and parameters optimization for reaction-diffusion models describing various natural phenomena can be found in [6-9].

Optimal control problems for the model of EVLA are considered in [10,11]. In [10], an optimal control problem for the reaction-diffusion model describing the EVLA procedure is posed, which consists in approximation of a given temperature profile at a certain point of the model domain. In [11], the similar as in [10] optimal control problem is studied. Here, the objective functional is taken such that its minimization allows one to reach the given temperature distribution in different parts of the model domain. This makes it possible to provide a sufficiently high temperature inside the vein for its successful obliteration and a safe temperature in the perivenous tissue. The unique solvability of the initial-boundary value problem is proved, on the basis of which the solvability of the optimal control problem is shown. An algorithm for finding a solution of the optimal control problem is proposed. Its efficiency is illustrated by a numerical example.

In the current work, an optimal control problem for quasilinear equations of radiative-conductive heat transfer simulating the process of endovenous laser ablation in a bounded domain $\Omega$ with the reflecting boundary $\Gamma=\partial \Omega$ is considered. The problem is to minimize the functional,
$$
J(\theta)=\int_{0}^{T} \int_{G_{1}}\left(\theta-\theta_{d}\right)^{2} d x d t \rightarrow \inf
$$
on the solutions of the initial-boundary value problem:
$$
\begin{aligned}
&\sigma \partial \theta / \partial t-\operatorname{div}(k(\theta) \nabla \theta)-\beta \varphi=u_{1} \chi \\
&-\operatorname{div}(\alpha \nabla \varphi)+\beta \varphi=u_{2} \chi, \quad x \in \Omega, \quad t \in(0, T), \\
&\theta=\left.0\right|_{\Gamma}, \quad \alpha \partial_{n} \varphi+\left.2^{-1} \varphi\right|_{\Gamma}=0,\left.\quad \theta\right|_{t=0}=\theta_{0}
\end{aligned}
$$
This takes into account the restrictions:
$$
u_{1,2} \geq 0, \quad u_{1}+u_{2} \leq P,\left.\quad \theta\right|_{G_{2}} \leq \theta_{*}
$$
Here, $G_{1}$ and $G_{2}$ are subdomains of $\Omega, \theta$ is the difference between the real temperature (units in Celsius) and the constant boundary temperature, $\varphi$ is the intensity of radiation averaged over all directions, $\alpha$ is the photon diffusion coefficient, $\beta$ is the absorption coefficient, $k(\theta)$ is the coefficient of thermal conductivity, $\sigma(x, t)$ is the product of the specific heat capacity and the volume density, $u_{1}$ describes the power of the source spending on heating the fiber tip, $u_{2}$ is the power of the source spending on radiation, $P$ is the maximal power of laser source, $\chi$ is the characteristic function of the part of the medium in which the fiber tip is located divided by the volume of the fiber tip. Note that the values of the parameters $u_{1}$ and $u_{2}$ are determined by a technique of the fiber tip carbonization. An overview of some ways of the carbonization is given in $[10]$.

Thus, when modeling the EVLA procedure, we will use the diffusion model that takes into account the conductive heat transfer, as well as the radiation transfer and absorption with the release of heat. The flow of bubbles formed at the heated optical fiber tip makes a significant contribution to the temperature field. In [2-4], based on an evaluation of experimental data, the heat transfer by the flow of bubbles is modeled using a piecewise constant thermal conductivity coefficient which depends on temperature as follows: when the temperature at some point reaches $95^{\circ} \mathrm{C}$, the coefficient of thermal conductivity increases 200 times.

Following the optimal control problem, it is required to provide a given distribution of the temperature field $\theta_{d}$ in the subdomain $G_{1}$, while the temperature in the subdomain $G_{2}$ cannot exceed (if it is possible) the critical value $\theta_{*}=$ Const $>0$. 2 FORMALIZATION OF THE OPTIMAL CONTROL PROBLEM

In what follows, we assume that $\Omega$ is a Lipschitz bounded domain, $\Gamma=\partial \Omega, Q=\Omega \times(0, T)$, $\Sigma=\Gamma \times(0, T)$. We denote by $L^{p}, 1 \leq p \leq \infty$, the Lebesgue space, by $H^{1}$ the Sobolev space $W_{2}^{1}$, by $H_{0}^{1}$ a subspace of functions from $H^{1}$ with zero boundary values, and by $L^{p}(0, T ; X)$ the Lebesgue space of functions from $L^{p}$, defined on $(0, T)$, with values in Banach space $X$.

Let $H=L^{2}(\Omega), V=H_{0}^{1}(\Omega)$, and the space $V^{\prime}$ be the dual to $V$. Then we identify $H$ with its dual space $H^{\prime}$ such that $V \subset H^{1}(\Omega) \subset H=H^{\prime} \subset$ $\left(H^{1}(\Omega)\right)^{\prime} \subset V^{\prime}$, and denote by $\|\cdot\|$ the norm in $H$, and by $(h, v)$ the value of functional $h \in V^{\prime}$ on the element $v \in V$ coinciding with the inner product in $H$ if $h \in H$. The inner product in $V$ is defined as $(u, v)_{V}=(\nabla u, \nabla v)$

We assume that the following conditions hold:

@@

where $\sigma_{i}, k_{i}, \alpha_{i}$, and $\beta_{i}$ are positive constants.

We define a nonlinear operator $A: V \rightarrow V^{\prime}$ and linear operator $B: H^{1}(\Omega) \rightarrow\left(H^{1}(\Omega)\right)^{\prime}$ using the following equalities valid for any $\theta, v \in V, \varphi, w \in$ $H^{1}(\Omega)$
$$
\begin{aligned}
&(A(\theta), v)=(k(\theta) \nabla \theta, \nabla v)=(\nabla h(\theta), \nabla v) \\
&(B \varphi, w)=(\alpha \nabla \varphi, \nabla w)+(\beta \varphi, w)+2^{-1} \int_{\Gamma} \varphi w d \Gamma
\end{aligned}
$$
where
$$
h(s)=\int_{0}^{s} k(r) d r .
$$
Definition 1. Let $u_{1,2} \in L^{2}(0, T)$. The pair of functions $\theta \in L^{2}(0, T ; V), \varphi \in L^{2}\left(0, T ; H^{1}(\Omega)\right)$ is $a$ weak solution of the problem (1), (2) if $\sigma \theta^{\prime} \in$ $L^{2}\left(0, T ; V^{\prime}\right)$ and

$\sigma \theta^{\prime}+A(\theta)-\beta \varphi=u_{1} \chi, \quad \theta(0)=\theta_{0}, \quad B \varphi=u_{2} \chi$ where $\theta^{\prime}=d \theta / d t$.

It follows from the Lax-Milgram lemma that for any function $g \in H$ there is a unique solution of equation $B \varphi=g$. Moreover, the inverse operator $B^{-1}: H \rightarrow H^{1}(\Omega)$ is continuous. Therefore, we can exclude the radiation intensity $\varphi=u_{2} B^{-1} \chi$ and formulate the optimal control problem as follows.

\section{Problem (P)}
$$
J(\theta)=\int_{0}^{T} \int_{G_{1}}\left(\theta-\theta_{d}\right)^{2} d x d t \rightarrow \inf
$$
$$
\begin{aligned}
& \sigma \theta^{\prime}+A(\theta)=u, \quad \theta(0)=\theta_{0},\left.\quad \theta\right|_{G_{2}} \leq \theta_{*}, \quad u \in U_{a d},
\end{aligned}
$$
where
$$
\begin{array}{r}
U_{a d}=\left\{u=u_{1} \chi+u_{2} \beta B^{-1} \chi: u_{1,2} \in L^{2}(0, T),\right. \\
\left.u_{1,2} \geq 0, u_{1}+u_{2} \leq P\right\}
\end{array}
$$

\section{PRELIMINARY RESULTS}
Let us consider the problem
$$
\sigma \theta^{\prime}+A(\theta)=f, \quad \theta(0)=\theta_{0} .
$$
The following lemma is true [11].

Lemma 1. Let conditions (c1)-(c3) hold and $f \in$ $L^{2}\left(0, T ; V^{\prime}\right)$. Then there is a solution of problem (3) such that $\theta \in L^{\infty}(0, T ; H)$ and the following estimates are valid:
$$
\begin{gathered}
\|\theta(t)\|^{2} \leq \frac{K}{\sigma_{0}} \exp \frac{\sigma_{2} t}{\sigma_{0}} \quad \text { a.e. on }(0, T) \\
\int_{0}^{T}\|\theta(t)\|_{V}^{2} d t \leq \frac{K}{k_{0}}\left(1+\frac{\sigma_{2} T}{\sigma_{0}} \exp \frac{\sigma_{2} T}{\sigma_{0}}\right)
\end{gathered}
$$
where $K=\sigma_{1}\left\|\theta_{0}\right\|^{2}+k_{0}^{-1}\|f\|_{L^{2}\left(0, T ; V^{\prime}\right)}^{2}$.

The following result is important for establishing the non-emptiness of the set of admissible controlstate pairs.

Lemma 2. Let conditions (c1)-(c3) hold, $f=0$, $\theta_{0} \leq \theta_{*}$ a.e. in $\Omega$, and $\theta$ be a solution to the problem (3). Then $\theta \leq \theta_{*}$ a.e. in $\Omega \times(0, T)$.

Proof. Multiplying, in the sense of the inner product in $H$, the first equation of $(3)$ by $v=\max \{\theta-$ $\left.\theta_{*}, 0\right\} \in L^{2}(0, T ; V)$, we obtain
$$
\left(\sigma v^{\prime}, v\right)+(k(\theta) \nabla v, \nabla v)=0 .
$$
Discarding the nonnegative second term, we arrive at the estimate
$$
\frac{d}{d t}(\sigma v, v) \leq\left(\sigma_{t} v, v\right) \leq \sigma_{2}\|v\|^{2} .
$$
Taking into account that $\left.v\right|_{t=0}=0$, integrate the last inequality over time. Then
$$
\sigma_{0}\|v(t)\|^{2} \leq(\sigma v(t), v(t)) \leq \sigma_{2} \int_{0}^{t}\|v(\tau)\|^{2} d \tau
$$
Based on Gronwall's lemma, we conclude that $v=0$ and therefore $\theta \leq \theta_{*}$ a.e. in $\Omega \times(0, T)$. 4 Solvability OF THE OPTIMAL CONTROL PROBLEM

Theorem 1. Let conditions (c1)-(c3) hold, and $\theta_{0} \leq \theta_{*}$ a.e. in $\Omega$. Then a solution to the problem $(\mathrm{P})$ exists.

Proof. By Lemmas 1 and 2, the set of admissible pairs is nonempty. Let us consider a minimizing sequence of admissible pairs $\left\{\theta_{m}, u_{m}\right\} \in$ $L^{2}(0, T ; V) \times U_{a d}$ such that $J\left(\theta_{m}\right) \rightarrow j=\inf J$ where
$$
\sigma \theta_{m}^{\prime}+A\left(\theta_{m}\right)=u_{m}, \quad \theta_{m}(0)=\theta_{0},\left.\quad \theta_{m}\right|_{G_{2}} \leq \theta_{*} .
$$
The boundedness in $L^{2}(0, T ; H)$ of the set of admissible controls $U_{a d}$ implies, by Lemma 1 , the estimates:
$$
\begin{gathered}
\left\|\theta_{m}\right\|_{L^{\infty}(0, T ; H)} \leq C, \quad\left\|\theta_{m}\right\|_{L^{2}(0, T ; V)} \leq C, \\
\left\|h\left(\theta_{m}\right)\right\|_{L^{2}(0, T ; V)} \leq C .
\end{gathered}
$$
Here and below in the proof of the theorem, by $C$ we denote constants independent on $m$. The estimates (5), using if necessary subsequences, lead to the existence of functions $u \in U_{a d}, \quad \theta \in L^{2}(0, T ; V)$, $\chi \in L^{2}(0, T ; V)$ such that
$$
\begin{aligned}
& u_{m} \rightarrow u \text { weakly in } L^{2}(0, T ; H), \\
& \theta_{m} \rightarrow \theta \text { weakly in } L^{2}(0, T ; V) \text {, } \\
& \text { *-weakly in } L^{\infty}(0, T ; H) \text {, } \\
& h\left(\theta_{m}\right) \rightarrow \chi \text { weakly in } L^{2}(0, T ; V) \text {. }
\end{aligned}
$$
The convergence results (6) are sufficient for passing to the limit as $m \rightarrow \infty$ in the system (4) and proving that the limit function $\theta \in L^{2}(0, T ; V)$ is such that $\sigma \theta^{\prime} \in L^{2}\left(0, T ; V^{\prime}\right)$ satisfies the equality
$$
\left(\sigma \theta^{\prime}, v\right)+(\nabla \chi, \nabla v)=(u, v) \quad \forall v \in V
$$
and the initial condition is valid.

The following estimate guarantees the compactness of the sequence $\theta_{m}$ in $L^{2}(Q)$ :
$$
\int_{0}^{T-\delta}\left\|\theta_{m}(s+\delta)-\theta_{m}(s)\right\|^{2} d s \leq C \delta
$$
From the inequality (7), using if necessary subsequences, we obtain that $\theta_{m} \rightarrow \theta$ in $L^{2}(Q)$. Therefore, by virtue of the inequality
$$
\left|h\left(\theta_{m}\right)-h(\theta)\right| \leq k_{1}\left|\theta_{m}-\theta\right|,
$$
we conclude that $h\left(\theta_{m}\right) \rightarrow h(\theta)$ in $L^{2}(Q)$ and hence $\chi=h(\theta)$. In addition, the limit function $\theta$ satisfies the inequality $\left.\theta\right|_{G_{2}} \leq \theta_{*}$. Hence, the limit pair $\{\theta, u\} \in L^{2}(0, T ; V) \times U_{a d}$ is admissible. Since the functional $J$ is weakly lower semicontinuous,
$$
j \leq J(\theta) \leq \liminf J\left(\theta_{m}\right)=j,
$$
then the pair $\{\theta, u\}$ is a solution of the problem $(\mathrm{P})$.

\section{PENALTY PROBLEM}
To solve numerically the optimal control problem with phase constraints $\left.\theta\right|_{G_{2}} \leq \theta_{*}$, consider the following penalty problem.

$\operatorname{Problem}\left(\mathbf{P}_{\varepsilon}\right): J_{\varepsilon}(\theta) \rightarrow \inf$, where
$$
\begin{aligned}
& J_{\varepsilon}(\theta)=\int_{0}^{T} \int_{G_{1}}\left(\theta-\theta_{d}\right)^{2} d x d t \\
& +\frac{1}{\varepsilon} \int_{0}^{T} \int_{G_{2}} F(\theta) d x d t, \\
& \sigma \theta^{\prime}+A(\theta)=u, \quad \theta(0)=\theta_{0}, \quad u \in U_{a d} .
\end{aligned}
$$
Here,
$$
F(\theta)= \begin{cases}0, & \text { if } \theta \leq \theta_{*} \\ \left(\theta-\theta_{*}\right)^{2}, & \text { if } \theta>\theta_{*}\end{cases}
$$
The estimates presented in Lemma 1 also make it possible to prove the solvability of the problem with the penalty similar to the proof of Theorem 1 .

>>>>

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[russian]{babel}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mhchem}
\usepackage{stmaryrd}
\usepackage{bbold}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\graphicspath{ {./images/} }

@@

\begin{abstract}
Исследуется задача оптимального управления для квазилинейных уравнений радиационно-кондуктивного
теплообмена, моделирующая процесс внутривенной лазерной
абляции в ограниченной области с отражающей границей.
В данной подобласти требуется обеспечить заданное
распределение температуры, а в другой
данной подобласти температура не может превышать критического значения.
На основе оценок решения управляемой системы
доказывается разрешимость задачи управления.
Предложен алгоритм
решения задачи оптимального управления путем минимизации целевого функционала со штрафом.
Работоспособность алгоритма иллюстрируется численным примером.
\end{abstract}

\section{Введение}
Процедура эндовенозной лазерной абляции (EVLA) безопасна и достаточно
эффективна при лечении варикозного расширения вен [1].
Во время ЭВЛА в поврежденную вену вводится лазерное оптическое волокно.
Затем лазерное излучение передается по волокну, которое в это время вытягивается из вены.
Конец оптического волокна обычно покрыт карбонизированным слоем (наконечник оптического волокна).
Карбонизированный слой разделяет лазерную энергию на нагрев наконечника оптического волокна и излучение.
Тепло от наконечника оптического волокна передается крови за счет кондуктивного теплообмена,
теплопередача значительно увеличивается за счет потока пузырьков,
образующихся на нагретом наконечнике волокна.
Излучение, попадающее в кровь и окружающие ткани, частично поглощается с выделением тепла.
В результате генерируемая тепловая энергия вызывает значительный нагрев вены,
что приводит к ее облитерации (закрытию вены).

Математическое моделирование радиационных и тепловых процессов, возникающих при ЭВЛА,
важно для определения оптимальных параметров излучения, обеспечивающих достаточно
высокую температуру внутри вены для успешной облитерации, с другой стороны,
генерируемое температурное поле должно быть относительно безопасным для тканей, окружающих вену.
Результаты численного моделирования EVLA для длин волн в диапазоне от
810 до $1470 \mathrm{~нм}$ и различных диаметров жил обсуждаются в [2-4].
В [5] отмечается важность математического моделирования для выбора оптимальной
мощности лазерного излучения и скорости отката волокна.
Обычно при выборе оптимальных параметров излучения
используется прямое множественное моделирование [2-5].
Такой подход неэффективен с практической точки зрения.
Более того, вопросы, связанные с существованием и
уникальностью оптимального режима, остаются открытыми.

Наиболее перспективным подходом к выбору оптимальных параметров излучения является
рассмотрение задачи оптимального управления для уравнений
реакционно-диффузионного типа, описывающих процедуру EVLA.
Различные подходы к анализу и оптимизации параметров реакционно-диффузионных моделей,
описывающих различные природные явления, можно найти в [6-9].

Задачи оптимального управления для модели EVLA рассмотрены в [10,11].
В [10] ставится задача оптимального управления для реакционно-диффузионной модели,
описывающей процедуру EVLA, которая заключается в аппроксимации заданного температурного
профиля в определенной точке модельной области.
В [11] изучается аналогичная [10] задача оптимального управления.
Здесь целевой функционал берется таким образом, что его минимизация позволяет достичь
заданного распределения температуры в разных частях модельной области.
Это позволяет обеспечить достаточно высокую температуру внутри вены для ее успешной
облитерации и безопасную температуру в перивенозной ткани.
Доказана однозначная разрешимость начально-краевой задачи,
на основе которой показана разрешимость задачи оптимального управления.
Предложен алгоритм нахождения решения задачи оптимального управления.
Его эффективность проиллюстрирована численным примером.

В настоящей работе рассматривается задача оптимального управления для квазилинейных
уравнений радиационно-кондуктивного теплообмена, моделирующих процесс эндовенозной
лазерной абляции в ограниченной области $\Omega$ с отражающей границей $\Gamma=\partial\Omega$.
Проблема заключается в том, чтобы свести к минимуму функционал

\[ J(\theta)=\int_{0}^{T} \int_{G_{1}}\left(\theta-\theta_{d}\right)^{2} d x d t \rightarrow \inf \]

на решениях начально-краевой задачи:

\[
\begin{aligned}
&\sigma \partial \theta / \partial t-\operatorname{div}(k(\theta)
\nabla \theta)-\beta \varphi=u_{1} \chi \\
&-\operatorname{div}(\alpha \nabla \varphi)+\beta \varphi=u_{2}
\chi, \quad x \in \Omega, \quad t \in(0, T), \\
&\theta=\left.0\right|_{\Gamma}, \quad \alpha \partial_{n}
\varphi+\left.2^{-1} \varphi\right|_{\Gamma}=0,\left.\quad \theta\right|_{t=0}=\theta_{0}
\end{aligned}
\]

При этом учитываются ограничения:

\[
u_{1,2} \geq 0, \quad u_{1}+u_{2} \leq P,\left.\quad \theta\right|_{G_{2}} \leq \theta_{*}
\]

Здесь $G_{1}$ и $G_{2}$ подмножества $\Omega, \theta$
представляет собой разницу между реальной температурой
(в единицах Цельсия) и постоянной граничной температурой,
$\varphi$ является ли интенсивность излучения усредненной по всем направлениям, $\alpha$
является коэффициентом диффузии фотонов, $\beta$ является коэффициентом поглощения,
$k(\theta)$ является коэффициентом теплопроводности, $\sigma(x, t)$
является произведением удельной теплоемкости и объемной плотности, $u_{1}$
описывает мощность источника, затрачиваемую на нагрев наконечника волокна, $u_{2}$
это мощность источника, расходуемая на излучение,
$P$ является максимальной мощностью лазерного источника, $\chi$
является характерной функцией той части среды, в которой волокно
наконечник расположен деленным на объем волоконного наконечника.
Обратите внимание, что значения параметров $u_{1}$ and $u_{2}$
определяются методом карбонизации кончика волокна.
Обзор некоторых способов карбонизации приведен в $[10]$.

Таким образом, при моделировании процедуры EVLA мы будем использовать диффузионную модель,
которая учитывает кондуктивный теплообмен, а также перенос излучения и поглощение с выделением тепла.
Поток пузырьков, образующихся на нагретом наконечнике оптического волокна, вносит значительный
вклад в температурное поле. В [2-4], основываясь на оценке экспериментальных данных,
теплопередача потоком пузырьков моделируется с использованием кусочно-постоянного
коэффициента теплопроводности, который зависит от температуры следующим образом:
когда температура в некоторой точке достигает $95 ^ {\circ} \mathrm{C}$,
коэффициент теплопроводность увеличивается в 200 раз.

Следуя задаче оптимального управления, требуется обеспечить заданное распределение
температурного поля $\theta_{d}$ в поддомене $G_{1}$,
при этом температура в поддомене $G_{2}$ не может превышать
(если это возможно) критическое значение $\theta_{*}=$ Const $>0$.

\textbf{2 ФОРМАЛИЗАЦИЯ ЗАДАЧИ ОПТИМАЛЬНОГО УПРАВЛЕНИЯ}

Будем далее предполагать, что $\Omega$ является липшицевой ограниченной областью,
$\Gamma=\partial \Omega, Q=\Omega \times(0, T)$, $\Sigma=\Gamma \times(0, T)$.
Обозначим через $L^{p}, 1 \leq p \leq \infty$, пространство Лебега,
через $H^{1}$ пространство Соболева $W_{2}^{1}$,
через $H_{ 0}^{1}$ подпространство функций из $H^{1}$ с нулевыми граничными значениями,
а через $L^{p}(0, T ; X)$ пространство Лебега функций из $L^{ p}$,
определенный на $(0, T)$, со значениями в банаховом пространстве $X$.

Пусть $H=L^{2}(\Omega), V=H_{0}^{1}(\Omega)$, а пространство $V^{\prime}$ двойственно к $V$.
Затем мы отождествляем $H$ с его двойственным пространством $H^{\prime}$ таким,
что $V \subset H^{1}(\Omega) \subset H=H^{\prime} \subset$
$\left( H^{1}(\Omega)\right)^{\prime} \subset V^{\prime}$,
и обозначим через $\|\cdot\|$ норму в $H$,
а через $(h , v)$ значение функционала $h \in V^{\prime}$
на элементе $v \in V$, совпадающее со скалярным произведением в $H$, если $h \in H$.
Скалярный продукт в $V$ определяется как $(u, v)_{V}=(\nabla u, \nabla v)$.

Будем предполагать, что выполнены следующие условия:

@@

где $\sigma_{i}, k_{i}, \alpha_{i}$, и $\beta_{i}$ положительные константы.

Определим нелинейный оператор $A: V \rightarrow V^{\prime}$ и линейный оператор
$B: H^{1}(\Omega) \rightarrow\left(H^{1}(\Omega)\right)^{\prime}$
используя следующие равенства, справедливые для любого
 $\theta, v \in V, \varphi, w \in$ $H^{1}(\Omega)$

\[
\begin{aligned}
&(A(\theta), v)=(k(\theta) \nabla \theta, \nabla v)=(\nabla h(\theta), \nabla v) \\
&(B \varphi, w)=(\alpha \nabla \varphi, \nabla w)+(\beta \varphi, w)+2^{-1}
\int_{\Gamma} \varphi w d \Gamma
\end{aligned}
\]

где
\[
h(s)=\int_{0}^{s} k(r) d r .
\]

\textbf{Definition 1.} Пусть $u_{1,2} \in L^{2}(0, T)$.
Пара функций $\theta \in L^{2}(0, T ; V), \varphi \in L^{2}\left(0, T ; H^{1}(\Omega)\right)$
тогда $a$ ялвяется слабым решением задачи (1), (2)
если $\sigma \theta^{\prime} \in$ $L^{2}\left(0, T ; V^{\prime}\right)$ и

$\sigma \theta^{\prime}+A(\theta)-\beta \varphi=u_{1} \chi, \quad \theta(0)=\theta_{0},
\quad B \varphi=u_{2} \chi$ where $\theta^{\prime}=d \theta / d t$.
Из леммы Лакса-Мильграма следует, что для любой функции $g \in H$ существует единственное
решение уравнения $B \varphi=g$.
Более того, обратный оператор $B^{-1}: H \rightarrow H^{1}(\Omega)$ непрерывен.
Поэтому можно исключить интенсивность излучения $\varphi=u_{2} B^{-1} \chi$ и
сформулировать задачу оптимального управления следующим образом.
\section{Problem (P)}
\[
J(\theta)=\int_{0}^{T} \int_{G_{1}}\left(\theta-\theta_{d}\right)^{2} d x d t \rightarrow \inf
\]

\[
\begin{aligned}
& \sigma \theta^{\prime}+A(\theta)=u, \quad \theta(0)=\theta_{0},\left.\quad
\theta\right|_{G_{2}} \leq \theta_{*}, \quad u \in U_{a d},
\end{aligned}
\]

где

\[
\begin{array}{r}
U_{a d}=\left\{u=u_{1} \chi+u_{2} \beta B^{-1} \chi: u_{1,2} \in L^{2}(0, T),\right. \\
\left.u_{1,2} \geq 0, u_{1}+u_{2} \leq P\right\}
\end{array}
\]

\section{ПРЕДВАРИТЕЛЬНЫЕ РЕЗУЛЬТАТЫ}
Давайте рассмотрим проблему
\[
\sigma \theta^{\prime}+A(\theta)=f, \quad \theta(0)=\theta_{0} .
\]
Справедлива следующая лемма [11].

\textit{Lemma 1.}
Пусть выполняются условия (c1)-(c3) и $f \in$ $L^{2}\left(0, T ; V^{\prime}\right)$.
Тогда существует решение задачи (3) такое, что $\theta \in L^{\infty}(0, T ; H)$
и справедливы следующие оценки:
\[
\begin{gathered}
\|\theta(t)\|^{2} \leq \frac{K}{\sigma_{0}} \exp \frac{\sigma_{2} t}{\sigma_{0}}
\quad \text { a.e. on }(0, T) \\
\int_{0}^{T}\|\theta(t)\|_{V}^{2} d t \leq
\frac{K}{k_{0}}\left(1+\frac{\sigma_{2} T}{\sigma_{0}} \exp
\frac{\sigma_{2} T}{\sigma_{0}}\right)
\end{gathered}
\]

где $K=\sigma_{1}\left\|\theta_{0}\right\|^{2}+k_{0}^{-1}\|f\|_{L^{2}\left(0, T ; V^{\prime}\right)}^{2}$.

Следующий результат важен для установления непустоты множества допустимых пар управляющих состояний.
\textit{Lemma 2.}Пусть выполняются условия (c1)-(c3) и
$f=0$, $\theta_{0} \leq \theta_{*}$ a.e. в $\Omega$, и $\theta$ решение задачи (3).
Тогда $\theta \leq \theta_{*}$ a.e. в $\Omega \times(0, T)$.

\textbf{Proof.}
Скалярно умножим в $H$ первое уравнение $(3)$ на
$v=\max \{\theta-$ $\left.\theta_{*}, 0\right\} \in L^{2}(0, T ; V)$ получаем

\[
\left(\sigma v^{\prime}, v\right)+(k(\theta) \nabla v, \nabla v)=0 .
\]

Отбрасывая неотрицательный второй член, приходим к оценке
\[
\frac{d}{d t}(\sigma v, v) \leq\left(\sigma_{t} v, v\right) \leq \sigma_{2}\|v\|^{2} .
\]
Учитывая, что $\left.v\right|_{t=0}=0$,проинтегрировать последнее неравенство по времени.
Тогда
\[
\sigma_{0}\|v(t)\|^{2} \leq(\sigma v(t), v(t)) \leq \sigma_{2} \int_{0}^{t}\|v(\tau)\|^{2} d \tau
\]

На основании леммы Гронуолла заключаем,
что $v=0$ и, следовательно, $\theta \leq \theta_{*}$ почти всюду в $\Omega\times(0,T)$.

\textbf{4. РАЗРЕШИМОСТЬ ЗАДАЧИ ОПТИМАЛЬНОГО УПРАВЛЕНИЯ}

\textbf{Theorem 1.}
Пусть выполняются условия (c1)-(c3), и $\theta_{0} \leq \theta_{*}$ a.e. в $\Omega$.
Тогда существует решение задачи $(\mathrm{P})$.

\textbf{Proof.}
Согласно леммам 1 и 2 множество допустимых пар непусто.
Рассмотрим минимизирующую последовательность допустимых
пар $\left\{\theta_{m}, u_{m}\right\} \in$ $L^{2}(0, T ; V) \times U_{a d}$
такой, что $J\left(\theta_{m}\right) \rightarrow j=\inf J$, где
\[
\sigma \theta_{m}^{\prime}+A\left(\theta_{m}\right)=u_{m},
\quad \theta_{m}(0)=\theta_{0},\left.\quad \theta_{m}\right|_{G_{2}} \leq \theta_{*} .
\]

Ограниченность в $L^{2}(0, T ; H)$ множества допустимых управлений $U_{a d}$ влечет по лемме 1 оценки:

\[
\begin{gathered}
\left\|\theta_{m}\right\|_{L^{\infty}(0, T ; H)} \leq C,
\quad\left\|\theta_{m}\right\|_{L^{2}(0, T ; V)} \leq C, \\
\left\|h\left(\theta_{m}\right)\right\|_{L^{2}(0, T ; V)} \leq C .
\end{gathered}
\]

Здесь и далее при доказательстве теоремы через $C$ обозначаются константы, не зависящие от $m$.
Оценки (5), используя при необходимости подпоследовательности,
приводят к существованию функций
$u \in U_{a d}, \quad \theta \in L^{2}(0, T ; V)$, $\chi \ в L^{2}(0, T ; V)$
такое, что

\[
\begin{aligned}
& u_{m} \rightarrow u \text { weakly in } L^{2}(0, T ; H), \\
& \theta_{m} \rightarrow \theta \text { weakly in } L^{2}(0, T ; V) \text {, } \\
& \text { *-weakly in } L^{\infty}(0, T ; H) \text {, } \\
& h\left(\theta_{m}\right) \rightarrow \chi \text { weakly in } L^{2}(0, T ; V) \text {. }
\end{aligned}
\]

Результаты сходимости (6) достаточны для предельного перехода
при $m \rightarrow \infty$ в системе (4) и доказательства того,
что предельная функция $\theta \in L^{2}(0, T ; V) $ такова,
что $\sigma \theta^{\prime} \in L^{2}\left(0, T ; V^{\prime}\right)$ удовлетворяет равенству

\[
\left(\sigma \theta^{\prime}, v\right)+(\nabla \chi, \nabla v)=(u, v) \quad \forall v \in V
\]

и начальное условие верно.

Следующая оценка гарантирует компактность последовательности $\theta_{m}$ in $L^{2}(Q)$:
\[
\int_{0}^{T-\delta}\left\|\theta_{m}(s+\delta)-\theta_{m}(s)\right\|^{2} d s \leq C \delta
\]

Из неравенства (7), используя при необходимости подпоследовательности, получаем, что
$\theta_{m} \rightarrow \theta$ in $L^{2}(Q)$.
Следовательно, в силу неравенства

\[
\left|h\left(\theta_{m}\right)-h(\theta)\right| \leq k_{1}\left|\theta_{m}-\theta\right|,
\]
следует, что $h\left(\theta_{m}\right) \rightarrow h(\theta)$ in $L^{2}(Q)$
и, следовательно $\chi=h(\theta)$.
Кроме того, предельная функция $\theta$ удовлетворяет неравенству
$\left.\theta\right|_{G_{2}} \leq \theta_{*}$.
Следовательно, допустима предельная пара $\{\theta, u\} \in L^{2}(0, T ; V) \times U_{a d}$.
Поскольку функционал $J$ слабо полунепрерывен снизу,

\[
j \leq J(\theta) \leq \liminf J\left(\theta_{m}\right)=j,
\]

тогда пара $\{\theta, u\}$ является решением задачи $(\mathrm{P})$.

\section{ЗАДАЧА ШТРАФОВ}

Чтобы численно решить задачу оптимального управления с фазовыми ограничениями
$\left.\theta\right|_{G_{2}} \leq \theta_{*}$, рассмотрим следующую задачу штрафов.
